The following section describes \emph{ejip}, a small TCP/IP stack
written entirely in Java for small, resource constraint devices.

%\section{Introduction}
%
%About ejip....

\section{TCP/IP Decisions}

\begin{itemize}
    \item Ignore IP options (TODO copy the data)
    \item Ignore Fragmentation
    \begin{itemize}
        \item Can we do this?
        \item Use and accept TCP max. length
        \item Restrict UDP/TCP to 512 bytes
        \item What is than the point for 1500 bytes Ethernet
        packages?
    \end{itemize}
    \item Should we go for simple routing?
    \item TCP/IP as stop \& go (like TFTP) -- but do a window probe
    (p 329)
    \begin{itemize}
        \item Sending: send segment and wait for ack
        \item Ignore out-of-order packets -- drop packets with not
        expected sequence \#
        \item Use the notion of TCPConnection
    \end{itemize}
    \item Don't bother with half close
    \item Do an event based API for TCP/IP and hide it by an
    optional Layer (Stream)
    \item Zero copy design: use the packet up to the application
    (e.g.\ TCP server)
    \item MSS at SYN (for MTU) -- use 512 byte data and check the
    link layer MTU (e.g.\ Slip, Ppp)
    \item Do the buffering at application level
    \begin{itemize}
        \item Really decided?
        \item Get data with sequence \#!
        \item Can be hidden by a stream as in jtcpip
    \end{itemize}
    \item There is no such thing as UDPConnection. UDP is
    connectionless!
\end{itemize}
\section{Notes}

\begin{itemize}
    \item Use smaller (and more) buffers and hold the buffer until
    ack or timeout.
    \item Keep the small web server (Html.java) as stateless (single) web
    page (probably with applet in code byte array)
    \item Interaction diagram would be helpful.
    \item Do timeouts on a simple 500~ms timer. Invoke
    System.currentTimeMillis() only once for the timer tick
    \item Check for minimal size (get rid of JDK code)
    \item Each part of the stack should implement Runable for the
    loop
    \item Event driven mode does not support blocking wait
    abstraction
    \item Prepare for performance measurements
    \begin{itemize}
        \item UDP (small/large) packages, multithreaded on PC
        \item TCP (small/large)
        \item Compare with jtcpip
        \item Where is the bottleneck (checksum, copy in jtcpip)
    \end{itemize}
    \item protothread switch expansion is very interesting for
    non-blocking (no wFNP()) profiles
\end{itemize}

Further (requested) properties of ejip:
\begin{itemize}
    \item Can run as single thread application -- don't use blocking
    (stop-and-wait) semantics of sockets -- fits even for SJC level
    0
    \item Invoking the application on a packet is event driven
    handling
    \item WCET analyzable! Check it with WCA and Volta
\end{itemize}

\subsection{Buffer Handling}

Should we use lists instead of linear search? Linear
    search is WCET predictable.

\begin{itemize}
    \item Application send:
    \begin{enumerate}
        \item Request a packet from the pool
        \item Fill it with data
        \item Return it to the pool for processing
    \end{enumerate}
    \item Interface receive:
    \begin{enumerate}
        \item Request a packet
        \item Fill it
        \item Return it to the pool for stack handling
    \end{enumerate}
    \item Interface send ready:
    \begin{enumerate}
        \item Grab a send read packet
        \item Send it
        \item Free the packet
    \end{enumerate}
    \item Grab a received packet:
    \begin{enumerate}
        \item Check IP header
        \item Invoke UDP/TCP/ICMP method
        \item Question: who frees the packet?
    \end{enumerate}
    \item Application receive invoked by the network code
\end{itemize}

\section{TCP/IP}

Two possibilities:
\begin{enumerate}
    \item TCP packets are kept in retransmission state till ack
    arrives
    \item The application has to reproduce the data (uip)
\end{enumerate}

The control is static (singleton), bit the data is OO.

\subsection{To Check}

\begin{itemize}
    \item Implement simple HTTP server with jtcpip
    \item Check uip -- done
    \item Check original TCP/IP source
\end{itemize}

\subsection{Start TCP/IP}

\begin{itemize}
    \item Start ejip documentation + do Javadoc
    \item Start with simple telnet/HTTP requests and get connection
    established working
    \begin{itemize}
        \item Basic done
        \item Retransmit a lost SYN missing
        \item Active open missing
    \end{itemize}
    \item Get a simple notion of time
    (Util/System.currentTimeMillis())
    \item on retransmission: add a new state for packet - sent, but
    not acked -- goes down to driver that send the data!
\end{itemize}

TCP/IP has to act on:
\begin{enumerate}
    \item Rcv package
    \item Write data from app
    \item Timeout (check connections)
\end{enumerate}

\subsection{TCP/IP Handler}

Event based and provides hooks for:
\begin{itemize}
    \item Connect
    \item Rcv data
    \item TX free?
    \item Close
    \item timeout?
\end{itemize}

Questions:
\begin{itemize}
    \item Is the application allowed to send data only when called
    form ejip or also independent?
    \item Application state in TCPConnection? or retransmission
    states?
    \item Stop/restart TCP/IP machine for application flow control?
\end{itemize}

\section{TODO}

\begin{itemize}
    \item Is checksum generation correct? (setting to 0xffff when 0,
    only in UDP?)
\end{itemize}
