\documentclass[a4paper,12pt]{scrartcl}
\usepackage{pslatex} % -- times instead of computer modern

\usepackage[colorlinks=true,linkcolor=black,citecolor=black]{hyperref}
\usepackage{booktabs}
\usepackage{graphicx}

\usepackage[latin1]{inputenc}

\usepackage{listings}
\lstset{basicstyle=\sffamily\small,keywordstyle=\sffamily\small,language=Java,captionpos=b,columns=flexible,showstringspaces=false}


\newcommand{\code}[1]{{\textsf{#1}}}
\newcommand{\sign}[1]{{\texttt{#1}}}
\newcommand{\scgrsc}{.8}
\newcommand{\scgrp}{.}
\hyphenation{SimpCon}

\begin{document}

\title{SimpCon -- a Simple SoC Interconnect\\Version 1.1}
\author{Martin Schoeberl\\ martin@jopdesign.com}
\maketitle \thispagestyle{empty}

\begin{abstract}
This document describes a simple interconnection standard for
system-on-chip (SoC) components. It is intended to provide pipelined
access to devices such on-chip peripherals and on-chip memory
controller with minimum hardware resources.
\end{abstract}

\subsection*{Versions}

\begin{description}
  \item[V 1.0] November 13, 2007, Initial version
  \item[V 1.1] June 1, 2009, Updated version
  \begin{itemize}
    \item Typo correction
    \item Additional signals for cache control
    \item A simple I/O example
  \end{itemize}
\end{description}

\input{simpcon_text}

\section{Summary}

This document describes a simple (with respect to the definition and
implementation) and efficient SoC interconnect \cite{simpcon}. The
novel signal \sign{rdy\_cnt} allows an early signalling to the master
when read data will be valid. This feature allows the master to
restart a stalled pipeline earlier to react for arriving data.
Furthermore, this feature also enables pipelined bus transactions
with a minimal effort on the master and the slave side.

We have compared SimpCon quantitative with AMBA and Avalon, two
common interconnection definitions. The application benchmark shows a
performance advantage of SimpCon by 17\% over AMBA and 19\% over
Avalon interfaces to an SRAM.

SimpCon is used as the main interconnect for the Java processor JOP
in a single master, multiple salves configuration. SimpCon is also
used to implement a shared memory chip-multiprocessor version of JOP.
Furthermore, in a research project on time-triggered network-on-chip
\cite{jop:ttnoc} SimpCon is used as the \emph{socket} to this NoC.

The author thanks Kevin Jennings and Tommy Thorn for the interesting
discussions about SimpCon, Avalon, and on-chip interconnection in
general at the Usenet newsgroup \texttt{comp.arch.fpga}.

\bibliographystyle{plain}
\bibliography{../../bib/all}

\end{document}
