% This is a copy from /usr/doc

\begin{itemize}

\subsubsection*{2003}

\item Martin Schoeberl.
 Using a {J}ava Optimized Processor in a Real World Application.
 In {\em Proceedings of the First Workshop on Intelligent Solutions in
  Embedded Systems (WISES 2003)}, pages 165--176, Austria, Vienna, June 2003.

\item Martin Schoeberl.
 Design Decisions for a {J}ava Processor.
 In {\em Tagungsband Austrochip 2003}, pages 115--118, Linz, Austria,
  October 2003.

\item Martin Schoeberl.
 {JOP}: {A} {J}ava Optimized Processor.
 In R.~Meersman, Z.~Tari, and D.~Schmidt, editors, {\em On the Move to
  Meaningful Internet Systems 2003: Workshop on {J}ava Technologies for
  Real-Time and Embedded Systems (JTRES 2003)}, volume 2889 of {\em Lecture
  Notes in Computer Science}, pages 346--359, Catania, Italy, November 2003.
  Springer.

\subsubsection*{2004}

\item Martin Schoeberl.
 Restrictions of {J}ava for Embedded Real-Time Systems.
 In {\em Proceedings of the 7th IEEE International Symposium on
  Object-Oriented Real-Time Distributed Computing (ISORC 2004)}, pages
  93--100, Vienna, Austria, May 2004.

\item Martin Schoeberl.
 Design Rationale of a Processor Architecture for Predictable
  Real-Time Execution of {J}ava Programs.
 In {\em Proceedings of the 10th International Conference on
  Real-Time and Embedded Computing Systems and Applications (RTCSA 2004)},
  Gothenburg, Sweden, August 2004.

\item Martin Schoeberl.
 Real-Time Scheduling on a {J}ava Processor.
 In {\em Proceedings of the 10th International Conference on
  Real-Time and Embedded Computing Systems and Applications (RTCSA 2004)},
  Gothenburg, Sweden, August 2004.

\item Martin Schoeberl.
 {J}ava Technology in an {FPGA}.
 In {\em Proceedings of the International Conference on
  Field-Programmable Logic and its applications (FPL 2004)}, Antwerp, Belgium,
  August 2004.

\item Martin Schoeberl.
 A Time Predictable Instruction Cache for a Java Processor.
 In Robert Meersman, Zahir Tari, and Angelo Corsario, editors, {\em On
  the Move to Meaningful Internet Systems 2004: Workshop on {J}ava Technologies
  for Real-Time and Embedded Systems (JTRES 2004)}, volume 3292 of {\em
  Lecture Notes in Computer Science}, pages 371--382, Agia Napa, Cyprus,
  October 2004. Springer.

\subsubsection*{2005}

\item Flavius Gruian, Per Andersson, Krzysztof Kuchcinski, and Martin Schoeberl.
 Automatic generation of application-specific systems based on a
  micro-programmed java core.
 In {\em Proceedings of the 20th ACM Symposium on Applied Computing,
  Embedded Systems track}, Santa Fee, New Mexico, March 2005.

\item Martin Schoeberl.
 Design and implementation of an efficient stack machine.
 In {\em Proceedings of the 12th IEEE Reconfigurable Architecture
  Workshop (RAW2005)}, Denver, Colorado, USA, April 2005. IEEE.

\item Martin Schoeberl.
 {\em JOP: A Java Optimized Processor for Embedded Real-Time Systems}.
 PhD thesis, Vienna University of Technology, 2005.

\item Martin Schoeberl.
 Evaluation of a {J}ava processor.
 In {\em Tagungsband Austrochip 2005}, pages 127--134, Vienna,
  Austria, October 2005.

\subsubsection*{2006}

\item Martin Schoeberl.
 A time predictable {J}ava processor.
 In {\em Proceedings of the Design, Automation and Test in Europe
  Conference (DATE 2006)}, pages 800--805, Munich, Germany, March 2006.

\item Martin Schoeberl.
 Real-time garbage collection for {J}ava.
 In {\em Proceedings of the 9th IEEE International Symposium on Object
  and Component-Oriented Real-Time Distributed Computing (ISORC 2006)}, pages
  424--432, Gyeongju, Korea, April 2006.

\item Martin Schoeberl. Instruction Cache f\"ur Echtzeitsysteme,
    April 2006. Austrian patent AT 500.858.

\item Rasmus Pedersen and Martin Schoeberl.
 An embedded support vector machine.
 In {\em Proceedings of the Fourth Workshop on Intelligent Solutions
  in Embedded Systems (WISES 2006)}, pages 79--89, Jun. 2006.

\item Rasmus Pedersen and Martin Schoeberl.
 Exact roots for a real-time garbage collector.
 In {\em Proceedings of the Workshop on {J}ava Technologies for
  Real-Time and Embedded Systems (JTRES 2006)}, Paris, France, October 2006.

\item Martin Schoeberl and Rasmus Pedersen.
 {WCET} analysis for a {Java} processor.
 In {\em Proceedings of the Workshop on {J}ava Technologies for
  Real-Time and Embedded Systems (JTRES 2006)}, Paris, France, October 2006.

\subsubsection*{2007}

\item Martin Schoeberl, Hans Sondergaard, Bent Thomsen, and Anders~P. Ravn.
A profile for safety critical java.
In {\em 10th IEEE International Symposium on Object and
  Component-Oriented Real-Time Distributed Computing (ISORC'07)}, pages
  94--101, Santorini Island, Greece, May 2007. IEEE Computer Society.

\item Martin Schoeberl.
Mission modes for safety critical java.
In {\em 5th IFIP Workshop on Software Technologies for Future
  Embedded \& Ubiquitous Systems}, May 2007.

\item Raimund Kirner and Martin Schoeberl.
Modeling the function cache for worst-case execution time analysis.
In {\em Proceedings of the 44rd Design Automation Conference, DAC
  2007}, San Diego, CA, USA, June 2007. ACM.

\item Martin Schoeberl.
A time-triggered network-on-chip.
In {\em International Conference on Field-Programmable Logic and its
  Applications (FPL 2007)}, Amsterdam, Netherlands, August 2007.

\item Christof Pitter and Martin Schoeberl.
Time predictable {CPU} and {DMA} shared memory access.
In {\em International Conference on Field-Programmable Logic and its
  Applications (FPL 2007)}, Amsterdam, Netherlands, August 2007.


\item Wolfgang Puffitsch and Martin Schoeberl.
{picoJava-II} in an {FPGA}.
In {\em Proceedings of the 5th international workshop on Java
  technologies for real-time and embedded systems (JTRES 2007)}, Vienna,
  Austria, September 2007. ACM Press.

\item Martin Schoeberl.
Architecture for object oriented programming languages.
In {\em Proceedings of the 5th international workshop on Java
  technologies for real-time and embedded systems (JTRES 2007)}, Vienna,
  Austria, September 2007. ACM Press.

\item Christof Pitter and Martin Schoeberl.
Towards a {Java} multiprocessor.
In {\em Proceedings of the 5th international workshop on Java
  technologies for real-time and embedded systems (JTRES 2007)}, Vienna,
  Austria, September 2007. ACM Press.

\item Martin Schoeberl and Jan Vitek.
Garbage collection for safety critical {Java}.
In {\em Proceedings of the 5th international workshop on Java
  technologies for real-time and embedded systems (JTRES 2007)}, Vienna,
  Austria, September 2007. ACM Press.

\item Martin Schoeberl. {SimpCon} - a simple and efficient {SoC}
    interconnect. In {\em Proceedings of the 15th Austrian
    Workhop on Microelectronics, Austrochip 2007}, Graz, Austria,
  October 2007.

\subsubsection*{2008}

\item Martin Schoeberl. A Java processor architecture for
    embedded real-time systems. {\em Journal of Systems
    Architecture}, 54/1--2:265--286, 2008.

\item Trevor Harmon, Martin Schoeberl, Raimund Kirner, and
    Raymond Klefstad. A modular worst-case execution time
    analysis tool for Java
  processors. In {\em Proceedings of the 14th IEEE Real-Time and
 Embedded Technology and Applications Symposium (RTAS 2008)}, St.
 Louis, MO, United States, April 2008.

\item Martin Schoeberl, Stephan Korsholm, Christian Thalinger,
    and Anders~P. Ravn. Hardware objects for {Java}. In {\em
    Proceedings of the 11th IEEE International Symposium on
  Object/component/service-oriented Real-time distributed
  Computing (ISORC 2008)}, Orlando, Florida, USA, May 2008. IEEE
  Computer Society.

\item Stephan Korsholm, Martin Schoeberl, and Anders~P. Ravn.
    Interrupt Handlers in {Java}.
 In {\em Proceedings of the 11th IEEE International Symposium on
  Object/component/service-oriented Real-time distributed
  Computing (ISORC 2008)}, Orlando, Florida, USA, May 2008. IEEE
  Computer Society.

\item Trevor Harmon, Martin Schoeberl, Raimund Kirner, and
    Raymond Klefstad. Toward libraries for real-time Java. In
    {\em Proceedings of the 11th IEEE International Symposium on
    Object/component/service-oriented Real-time distributed
  Computing (ISORC 2008)}, Orlando, Florida, USA, May 2008. IEEE
  Computer Society.

\item Christof Pitter and Martin Schoeberl. Performance
    evaluation of a {Java} chip-multiprocessor. In {\em
    Proceedings of the 3rd IEEE Symposium on Industrial Embedded
    Systems (SIES 2008)}, Jun. 2008.

\item Martin Schoeberl. Application experiences with a real-time
    {J}ava processor. In {\em Proceedings of the 17th IFAC World
    Congress}, Seoul, Korea, July 2008.

\item Peter Puschner and Martin Schoeberl. On composable system
    timing, task timing, and WCET analysis. In {\em Proceedings
    of the 8th International Workshop on Worst-Case Execution
    Time (WCET) Analysis}, Prague, Czech
  Republic, July 2008.

\item Martin Schoeberl. {\em JOP: A Java Optimized Processor for
    Embedded Real-Time Systems}. Number ISBN 978-3-8364-8086-4.
    VDM Verlag Dr. M{\"u}ller, July 2008.

\item Martin Schoeberl and Wolfgang Puffitsch. Non-blocking
    object copy for real-time garbage collection. In {\em
    Proceedings of the 6th International Workshop on Java
  Technologies for Real-time and Embedded Systems (JTRES 2008)},
  September 2008.

\item Wolfgang Puffitsch and Martin Schoeberl. Non-blocking root
    scanning for real-time garbage collection. In {\em
    Proceedings of the 6th International Workshop on Java
  Technologies for Real-time and Embedded Systems (JTRES 2008)},
  September 2008.

\item Walter Binder, Martin Schoeberl, Philippe Moret, and Alex
    Villazon. Cross-profiling for embedded Java processors. In
    {\em Proceedings of the 5th International Conference on the
    Quantitative Evaluation of SysTems (QEST 2008)}, St Malo,
  France, September 2008.

\item Walter Binder, Alex Villazon, Martin Schoeberl, and
    Philippe Moret. Cache-aware cross-profiling for Java
    processors. In {\em Proceedings of the 2008 international
    conference on Compilers, architecture, and synthesis
    forembedded systems (CASES 2008)}, Atlanta, Georgia, October
    2008. ACM.

\subsubsection*{2009}

\item Martin Schoeberl. Time-predictable computer architecture.
    {\em EURASIP Journal on Embedded Systems}, vol. 2009, Article
    ID 758480:17 pages, 2009.

\item Martin Schoeberl. Time-predictable cache organization. In
    {\em Proceedings of the First International Workshop on
    Software Technologies for Future Dependable Distributed
    Systems (STFSSD 2009)}, Tokyo, Japan, March
  2009. IEEE Computer Society.

\item Andy Wellings and Martin Schoeberl. Thread-local scope
    caching for real-time {J}ava. In {\em Proceedings of the 12th
    IEEE International Symposium on
  Object/component/service-oriented Real-time distributed
  Computing (ISORC 2009)}, Tokyo, Japan, March 2009. IEEE
  Computer Society.

\item Florian Brandner, Tommy Thorn, and Martin Schoeberl.
    Embedded {JIT} compilation with {CACAO} on {YARI}. In {\em
    Proceedings of the 12th IEEE International Symposium on
    Object/component/service-oriented Real-time
  distributed Computing (ISORC 2009)}, Tokyo, Japan, March 2009.
  IEEE Computer Society.

\item Thomas Henties, James~J. Hunt, Doug Locke, Kelvin Nilsen,
    Martin Schoeberl, and Jan Vitek. Java for safety-critical
 applications. In {\em 2nd International Workshop on the
 Certification of Safety-Critical Software Controlled Systems
 (SafeCert 2009)}, Mar. 2009.

\item Martin Schoeberl and Peter Puschner. Is
    chip-multiprocessing the end of real-time scheduling? In {\em
    Proceedings of the 9th International Workshop on Worst-Case
  Execution Time (WCET) Analysis}, Dublin, Ireland, July 2009.
  OCG.

\item Benedikt Huber and Martin Schoeberl. Comparison of implicit
    path enumeration and model checking based WCET
  analysis. In {\em Proceedings of the 9th International Workshop
on Worst-Case
  Execution Time (WCET) Analysis}, Dublin, Ireland, July 2009.
  OCG.

\item Philippe Moret, Walter Binder, Martin Schoeberl, Alex
    Villazon, and Danilo Ansaloni. Analyzing performance and
dynamic behavior of embedded Java software with calling-context
  cross-profiling. In {\em Proceedings of
the 7th International Conference on the Principles and Practice
  of Programming in Java (PPPJ 2009)}, Calgary, Alberta, Canada,
  August 2009. ACM.

\item Martin Schoeberl, Walter Binder, Philippe Moret, and Alex
    Villazon. Design space exploration
for Java processors with cross-profiling. In {\em Proceedings of
the 6th International Conference on the Quantitative Evaluation
  of SysTems (QEST 2009)}, Budapest, Hungary, September 2009.
  IEEE Computer Society.

\item Philippe Moret, Walter Binder, Alex Villazon, Danilo
    Ansaloni, and Martin Schoeberl. Locating performance
    bottlenecks in embedded Java software with
  calling-context cross-profiling. In {\em Proceedings of the 6th
International Conference on the
  Quantitative Evaluation of SysTems (QEST 2009)}, Budapest,
  Hungary, September 2009. IEEE Computer Society.

\item Jack Whitham, Neil Audsley, and Martin Schoeberl. Using
    hardware methods to improve time-predictable performance in
  real-time java systems. In {\em Proceedings of the 7th
International Workshop on Java
  Technologies for Real-time and Embedded Systems (JTRES 2009)},
  Madrid, Spain, September 2009. ACM Press.


\subsubsection*{2010}

\item Martin Schoeberl and Wolfgang Puffitsch. Non-blocking
    real-time garbage collection. {\em Trans. on Embedded
    Computing Sys.}, accepted, 2010.

\item Christof Pitter and Martin Schoeberl. A real-time {Java}
    chip-multiprocessor. {\em Trans. on Embedded Computing Sys.},
    accepted, 2010.

\item Walter Binder, Martin Schoeberl, Philippe Moret, and Alex
    Villazon. Cross-profiling for Java processors. {\em Software:
Practice and Experience}, accepted, 2010.

\end{itemize}
